Visualizing Contingency Tables

### 2006-08-23 trunk/vcd/paper/Shadings/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@551 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

data("Arthritis")
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")

data("UCBAdmissions")
ucb <- aperm(UCBAdmissions)

myseed <- 1071
@

<<inference,echo=FALSE,results=hide>>=
if(FALSE) {
  set.seed(myseed)
  art_max <- coindep_test(art, n = 5000)
  set.seed(myseed)
  art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))

  set.seed(myseed)
  art_gp <- shading_max(art, n = 5000)

  set.seed(myseed)
  ucb_doublemax <- coindep_test(ucb, "Dept", n = 5000)
  set.seed(myseed)
  ucb_maxChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  ucb_sumChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)

  set.seed(myseed)
  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000, margins = c(3, 1, 1, 3))

  save(art_max, art_chisq, art_gp,
       ucb_doublemax, ucb_maxChisq, ucb_sumChisq, ucb_cotab,
       file = "JCGS-inference.rda")
} else {
  load("JCGS-inference.rda")
}
@

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the appearance of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular conditional independence
problems.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing them out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing independence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
visualization and significance
testing. Friendly's shading is typically implemented in statistical software using color
spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
to properly read and interpret the corresponding plots. For overcoming this problem,
a general strategy for constructing diverging palettes in the perceptually-based HCL
space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
model of a contingency table more tightly than in previous approaches, the residual-based
shading is extended such that appearance of color in the display is equivalent to significance
of the associated independence test. This is achieved by using data-driven cut offs
for the appearance of color computed from the conditional permutation distribution
\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
residual-based shadings are illustrated both for mosaic and association plots 
using real world data sets.

@@

\begin{tabular}{rr|rrr}
\\
                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
                   &         & None & Some & Marked \\ \hline
\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
                   & Treated &    6 &    5 &     16
\end{tabular}

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks when addressing the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

However, the sum of squares is not the only plausible way of capturing deviations
from zero in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the `dependence' (i.e., significant
departure from independence), if any, is the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the null hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equation~\ref{eq:ChiSquare} was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~\ref{eq:H0}, using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $5000$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

Two well-established visualization techniques for independence in 2-way tables are
mosaic plots and association plots. Both are suitable to bring out
departures of an observed table $[n_{ij}]$ from the expected table $[\hat n_{ij}]$
in a graphical way. The latter focuses on the visualization of the Pearson residuals
$r_{ij}$ (under independence) while the former primarily displays the observed frequencies
$n_{ij}$.

@@

\emph{Association plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
same interpretation as the mosaic plot: there are more placebo patients with no 
improvement and fewer with marked improvement than expected under independence---vice
versa for the treated patiens.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
The implementation in the standard packages of the
\proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can generate the
same colors, we only discuss HSV space in the following.

The HSV space originally looks like a cone\footnote{The wikipedia gives a very useful
overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
links to comparisons with other color spaces discussed in this paper.}
with black at its peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
unit interval are used in this space, giving it the appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
the saturation/value plane for the given hues $h = 0$ and $h = 2/3$. 
The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in Section~\ref{sec:significance}.

@@

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

@@

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually-based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue-Chroma-Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV colors.

Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space and consequently cannot be standardized in the same way as the
HSV space. Its dimensions are usually given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent. However, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. For the task of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
$h = 260$ is chosen (rather than $h = 240$ which would correspond to $h = 2/3$ in HSV space)
because its chroma/luminance plane is most similar to that of $h = 0$.}
To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or un-interstingness, respectively). That is,
viewers might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. However,
in this situation, it would be nice if such cut offs could be chosen automatically in a
data-driven way. Strategies for this are derived in the following.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

>>>>

data("Arthritis", package = "vcd")
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
names(dimnames(art))[2] <- "Improvement"

data("pistonrings", package = "HSAUR")
pring <- pistonrings
names(dimnames(pring)) <- c("Compressor", "Leg")

data("alzheimer", package = "coin")
alz <- xtabs(~ smoking + disease + gender, data = alzheimer)
dimnames(alz)[[2]] <- c("Alzheimer's", "Dementia", "Other")
names(dimnames(alz)) <- c("Smoking", "Disease", "Gender")

data("Punishment", package = "vcd")
pun <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
dimnames(pun) <- list(Memory = c("yes", "no"), Attitude = c("no", "moderate"),
  Age = c("15-24", "25-39", "40+"), Education = c("Elementary", "Secondary", "High"))

myseed <- 1071
nrep <- 5000
recompute <- FALSE
@

<<inference,echo=FALSE,results=hide>>=
if(recompute) {
  set.seed(myseed)
  art_max <- coindep_test(art, n = nrep)
  set.seed(myseed)
  art_chisq <- coindep_test(art, n = nrep, indepfun = function(x) sum(x^2))

  set.seed(myseed)
  pring_max <- coindep_test(pistonrings, n = nrep)
  set.seed(myseed)
  pring_chisq <- coindep_test(pistonrings, n = nrep, indepfun = function(x) sum(x^2))

  set.seed(myseed)
  alz_dmax <- coindep_test(alz, "Gender", n = nrep)
  set.seed(myseed)
  alz_maxchisq <- coindep_test(alz, "Gender", n = nrep, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  alz_sumchisq <- coindep_test(alz, "Gender", n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
  set.seed(myseed)
  alz_cotab <- cotab_coindep(alz, condvars = "Gender", n = nrep, margins = c(2, 1, 1, 2),
    varnames = FALSE)

  set.seed(myseed)
  pun_dmax <- coindep_test(pun, 3:4, n = nrep)
  set.seed(myseed)
  pun_maxchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2))
  set.seed(myseed)
  pun_sumchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
  set.seed(myseed)
  pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
    varnames = FALSE, margins = c(2, 1, 1, 2))

  save(art_max, art_chisq,
       pring_max, pring_chisq,
       alz_dmax, alz_maxchisq, alz_sumchisq, alz_cotab,
       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab,
       file = "JCGS-inference.rda")
} else {
  load("JCGS-inference.rda")
}
@

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the appearance of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular conditional independence
models.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing them out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing dependence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
visualization and significance testing. Friendly's shading is typically implemented
in statistical software using color spaces such as HLS (Hue-Luminance-Saturation) or
HSV (Hue-Saturation-Value) colors. The dimensions of both spaces are only
poorly mapped to the perceptual dimensions of the
human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
to properly read and interpret the corresponding plots. For overcoming this problem,
a general strategy for constructing diverging palettes in the perceptually-based HCL
space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
model of a contingency table more tightly than in previous approaches, the residual-based
shading is extended such that appearance of color in the display is equivalent to significance
of the associated independence test. This is achieved by using data-driven cut offs
for the appearance of color computed from the conditional permutation distribution
\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
residual-based shadings are illustrated both for mosaic and association plots 
using real world data sets.

@@

\begin{tabular}{rr|rrr}
\\
                   &         & \multicolumn{3}{|l}{\textbf{Improvement}} \\
                   &         & None & Some & Marked \\ \hline
\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
                   & Treated &    6 &    5 &     16
\end{tabular}

@@

The estimated expected cell frequencies under $H_0$ are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their sum of squares
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks when addressing the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

However, the sum of squares is not the only plausible way of capturing deviations
from zero in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the `dependence' (i.e., significant
departure from independence), if any, is the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute values exceed $c_\alpha$ violate the null hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equation~\ref{eq:ChiSquare} was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests, conditioning on the observations)---carried
out either by simulation or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~\ref{eq:H0}, using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $\Sexpr{nrep}$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

Two well-established visualization techniques for independence in 2-way tables are
mosaic plots and association plots. Both are suitable to bring out
departures of an observed table $[n_{ij}]$ from the estimated expected table $[\hat n_{ij}]$
in a graphical way. The latter focuses on the visualization of the Pearson residuals
$r_{ij}$ (under independence) while the former primarily displays the observed frequencies
$n_{ij}$.

@@

\emph{Association plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
same interpretation as the mosaic plot: there are more placebo patients with no 
improvement and fewer with marked improvement than expected under independence---vice
versa for the treated patiens.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS color space.
The implementation in the standard packages of the
\proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can easily generate the
same colors, we only discuss HSV space in the following.

The HSV space looks like a cone\footnote{The wikipedia gives a very useful
overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
links to comparisons with other color spaces discussed in this paper.}
with black at its peak (zero value) and full color wheels for different saturations at the other end,
around a white center (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
unit interval are used in this space, giving it the appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
the saturation/value plane for the given hues $h = 0$ and $h = 2/3$. 
The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in Section~\ref{sec:significance}.

@@

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(50, 75, 100))
axis(4, at = c(50, 75, 100)/100, labels = c(50, 75, 100))
axis(3, at = -4:4*.25, labels=c(4:0*25, 1:4*25))
mtext(c("hue = 0", "hue = 240"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

@@

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually-based colors that are `in harmony' goes back until at least \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue-Chroma-Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations that the HSV colors have.

Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space and consequently cannot be standardized in the same way as the
HSV space. Its dimensions are usually given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent. However, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. For the task of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
$h = 260$ is chosen rather than $h = 240$ because its chroma/luminance plane is most similar,
as assessed by the symmetric difference of the planes, to that of $h = 0$.}
To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or un-interstingness, respectively). That is,
viewers might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both are not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence (as according
to both the maximum and $\chi^2$ tests), no residual exceeds an absolute value of $2$
and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. However,
in this situation, it would be nice if such cut offs could be chosen automatically in a
data-driven way. Strategies for this are derived in the following.

@@

\setkeys{Gin}{width=2\textwidth}
\begin{figure}[t!]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=12,width=16,results=hide,include=FALSE>>=
mymar <- c(1.5, 0.5, 0.5, 2.5)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 3)))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = mymar,
  gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
mosaic(art, newpage = FALSE, gp = shading_hsv, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 3))
mosaic(art, newpage = FALSE, gp = shading_hcl, margins = mymar,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()
